\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2009-12-03 honors_thesis/thesis/paper.tex
I'm pretty much done with the GECCO paper.
I significantly changed the results section, I added a number of
figures using xfig that really help out the explanation, and I
split the second distributed method into its own subsection.

\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has, because there are two possibilities for updating a
particle's personal best (updated and not updated) and $n+1$ possibilities for
updating its neighborhood best (one of $n$ neighbors, or no update).  If all of
those speculative positions are evaluated in parallel along with the original,
non-speculative position of the particle at time $t-1$, then one of the
speculative positions has to be the correct one, and two iterations have been
evaluated at once.

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children.
$\n_t$ is a neighbor of particle $\p_t$.  A set of particles is $\pset$,
$\sset$, or $\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.

@@

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with.  Each of those
combinations defines a speculative particle, $\s_t$, that can be moved using
the motion equations to obtain $\noeval{\s}_{t+1}$.  The result is a set of
particles $\noeval{\pset}_t$, and for each particle a set of speculative
children $\noeval{\sset}_{t+1}$, which can all be evaluated at once.

@@

This can be accomplished in two ways, either by performing several rounds of
very simple message passing, or by performing a single round with a lot more
messages.

The first method takes three rounds of passing information, two for iteration
$t$ and one for iteration $t+1$.  First, each particle must be moved from
iteration $t-1$ to $\noeval{\p}_t$.  Then a round of message passing is
required so that each particle knows the position of its neighbors at iteration
$t$, from which $\noeval{\sset}_{t+1}$ can be generated.  After each particle
evaluates its position and the positions of its speculative children, it passes
information about the outcome iteration $t$ to its neighbors, so that each
particle can receive information about its neighborhood best to be updated to
$\p_t$.  Once that communication is finished, the particle can select the
speculative child which matched the branch that iteration $t$ actually
produced.  Then another round of information passing follows, for iteration
$t+1$, so that $\nonbest{\p}_{t+1}$ can be updated to $\p_{t+1}$.  Two
iterations have then been completed with only one round of evaluations, and the
next iteration can start again with the first round of message passing.

Rounds of message passing can be expensive.  Accordingly, we describe a method
that requires only one round of communication for each pair of iterations.  A
lot more messages are needed, but that is sometimes more desirable than
synchronizing all of the machines three times.

@@

Upon evaluation, $\noeval{\p}_t$ becomes $\nonbest{\p}_t$, needing only to get
its neighborhood best information from its neighbors.  All of its neighbors,
then, must send it a message, so that from their updated personal best at
iteration $t$ the particle becomes $\p_t$.

With $\p_t$ we can select the correct speculative child as described above and
produce $\nonbest{\p}_{t+1}$.  We then need the set of neighbors to
$\nonbest{\p}_{t+1}$, $\nonbest{\nset}_{t+1}$, so we can update
$\nonbest{\p}_{t+1}$'s neighborhood best.

To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same information for
the neighboring particle that we needed to produce the original particle,
$\nonbest{\p}_{t+1}$; we need the original neighbor particle, its speculative
children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  With $\nonbest{\nset}_{t+1}$ and
$\nonbest{\p}_{t+1}$, we can produce $\p_{t+1}$.

In order to get $\p_{t+1}$, then, a particle needs to receive messages from its
neighbors, its neighbors' neighbors, its speculative children, and its
neighbors' speculative children.  \fig{origparticle} shows graphically how this
information is used to produce $\p_{t+1}$.  The particle $\p_{t+1}$ can be
passed to some central machine to track the progress of the algorithm, and it
can be moved to $\noeval{\p}_{t+2}$ in order to start the next iteration.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{messages.eps}
  \caption{The messages needed to produce $\p_{t+1}$. $nn$ are the neighbors
  of $p$'s neighbors, and $ns$ are the speculative children of $n$.  Items
  without an arrow pointing to them are messages the particle needs to receive.}
  \label{fig:origparticle}
\end{figure}

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.  
\fig{specparticles} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

\begin{figure}
  \centering
  \includegraphics[width=.6\columnwidth]{messages2.eps}
  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
  and $\nset_{t+1}$ are both produced as in \fig{origparticle}.}
  \label{fig:specparticles}
\end{figure}

@@

The decision between these two methods of performing speculative evaluation in
a distributed parallel PSO algorithm depends on which is more expensive: 
individual messages between processors or rounds of communication.

@@

The original PSO algorithm was developed with only static topologies in mind. A
few of the details about the neighborhood best $\nbest$ were left unspecified,
as they are not issues with static topologies.  However, when neighbors change
from iteration to iteration a few questions arise.  Is $\nbest$ the best
position of any of its all-time neighbors, or only its current neighbors?  Do
particles share their current value, their personal best value, or their
neighborhood best value to contribute to their neighbors' neighborhood best?

It has been shown that in dynamic topologies using the $\nbest$ only from
current neighbors leads to poor performance.  $\nbest$ should instead be
defined as the best position of any of the neighbors the particle has ever
had~\cite{mcnabb-cec09}.  The question of which value to share with neighbors
is an open one.  Giving more information to a particle's neighbors shares
information more quickly through the swarm.  We choose to share the personal
best value as a good middle ground.  Also, sharing the neighborhood best would
require more message passing in distributed algorithms, as particles would need
more information in order to generate their speculative children from their
neighbors.

Performing speculative evaluation in PSO with a dynamic topology also raises a
few issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random.

@@

This can be thought of as drawing a number of samples from the next iteration
and accepting the best one.  Speculative particles that move in good directions
are kept.  Clearly this modification to PSO would produce more exploitation,
decreasing the amount of exploration.  With functions that are less deceptive
we would expect this method to work well, while performance on more deceptive
functions would probably be hurt.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  To have the same number of function evaluations per iteration
it is necessary to use a smaller swarm size in the speculative algorithm.
There are $2(n+1)$ speculative evaluations per particle for every two
iterations, plus one for the original, non-speculative particle.  For each of
the topologies we used, a particle has three neighbors, including itself.  One
of the $2(n+1)$ cases can be eliminated, however, as the particle cannot be its
own neighborhood best update while not also updating its personal best
(otherwise its personal best would be the neighborhood best, and it would not
have updated).  So we have $2(n+1)$ total evaluations per particle, giving us
$8p_s$ evaluations for every two iterations, where $p_s$ is the number of
particles in the speculative swarm.  A regular swarm uses $2p_r$ evaluations
per two iterations.  To have the same number of function evaluations per
iteration, we have that $p_s = \frac{1}{4}p_r$.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used.  To be fair in our comparisons, for the standard PSO algorithm
we compare to both the Random topology with two neighbors and to the Complete
topology.  All functions have 20 dimensions, and all results shown are averages
over 20 runs.  Except where noted, we use swarms of close to 50 particles, as
that has been recommended as a general-purpose swarm size~\cite{bratton-sis07}.
When the speculative swarms have 50 particles, regular PSO swarms have 200.
For ease of labeling, we call our speculative algorithm SEPSO, the variant
SEPSO Pick Best, and standard PSO just PSO, with the topology following if
necessary.

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
that SEPSO clearly beat PSO with a Random topology.  However, the performance
between SEPSO and PSO Complete is not as encouraging---the results are
indistinguishable.  But SEPSO Pick Best handily outperforms all other
algorithms in this problem.

\begin{figure}
  \includegraphics[width=.8\columnwidth]{sphere1.eps}
  \caption{Function Sphere with a swarm that performs 104 evaluations per two
  iterations.  Speculative algorithms have a swarm size of 13, and the original
  PSO has a swarm size of 52.}
  \label{fig:sphere1}
\end{figure}

To further illustrate the point that performing speculative evaluation is a
much better idea with some functions than simply adding particles, in
\fig{sphere2} we show a graph comparing the performance of SEPSO Pick Best with
13 particles and with 50 particles.  The difference between the two in terms of
function evaluations is staggering.  However, this result is well known; with
Sphere and similar functions, the best thing to do is have a small swarm run
longer.  We show the results here because our method allows the algorithm to
use spare processors in parallel to perform more iterations at the same time.
Clearly with these functions the best thing to do is not to add particles to
the swarm, but to perform speculative evaluation.

\begin{figure}
  \includegraphics[width=.8\columnwidth]{sphere2.eps}
  \caption{Function Sphere with SEPSO Pick Best in two different swarm sizes.
  The difference in performance is incredible and suggests that for this
  function processors should most certainly be used to increase the number of
  iterations performed and not add more particles to the swarm.}
  \label{fig:sphere2}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for a swarm size of 50
particles with SEPSO.  One can see in the figure that both SEPSO algorithms
initially outperformed PSO, but got stuck in a local optima and were eventually
outperformed.

\begin{figure}
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that performs 400 evaluations per two
  iterations.  Speculative algorithms have a swarm size of 50, and the original
  PSO has a swarm size of 200.}
  \label{fig:griewank1}
\end{figure}

It has also been shown that with Griewank, swarms that pass a threshold of size
consistently find the global optimum.  That size is about 400
particles~\cite{mcnabb-cec09}.  Thus we also show results for a swarm size of
400 with SEPSO in \fig{griewank2}.  Once the speculative swarm is of a
sufficient size to not get caught in the local optima, it clearly outperforms
standard PSO, finding the global optimum in fewer function evaluations.

\begin{figure}
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that performs 3200 evaluations per
  two iterations.  Speculative algorithms have a swarm size of 400, and the
  original PSO has a swarm size of 1600.  The local optima with the lowest
  function value is close to $10^{-6}$.  Thus both algorithms consistently find
  the global optimum without getting stuck.}
  \label{fig:griewank2}
\end{figure}

Lastly we look at functions that do not show much or any improvement with this
method.  The Ackley function is defined as $f(\Vec{x}) = -20\exp\left\{
-0.2\sqrt{\frac{1}{D}\sum_{i=1}^Dx_i^2}\right\} - \exp\left\{\frac{1}{D}
\sum_{i=1}^D\cos(2\pi x_i)\right\}+20+e$. \fig{ackley} shows a comparison of
our speculative algorithms versus the original PSO with swarm sizes of 50 and
200, respectively.  There is no significant difference between the performance
of SEPSO and PSO Complete, though the performance of SEPSO Pick Best is
interesting to note.  It quickly converged on a local optimum and compared to
the other algorithms had dismal performance on this function.

We also ran experiments with the function Rastrigin, $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$,  though space does not
permit us to present a graph of our results.  However, performance of SEPSO and
PSO Complete was largely indistinguishable, as it was with Ackley.

\begin{figure}
  \includegraphics[width=.8\columnwidth]{ackley.eps}
  \caption{Function Ackley with a swarm that performs 400 evaluations per
  two iterations.  Speculative algorithms have a swarm size of 50, and the
  original PSO has a swarm size of 200.  Note the particularly horrible
  performance of SEPSO Pick Best.}
  \label{fig:ackley}
\end{figure}

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using this
algorithm, the original PSO algorithm is exactly reproduced two iterations at a
time.  This technique requires more function evaluations per iteration than
regular PSO, but for a number of functions still performs better.

We have also described a method for making use of extra speculative evaluations
that performs incredibly well on some functions and incredibly poorly on other
functions.

@@

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration, and it has been shown that for a 20-dimensional Griewank function,
400 particles is ``enough'' exploration~\cite{mcnabb-cec09}.  For these
functions, one should use processors to increase the swarm size until enough
exploration has been reached, then use additional processors to increase the
number of iterations performed.

>>>>

\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{psfrag}

@@

%SpecExPSO Stuff
\providecommand{\noeval}[1]{\ensuremath{#1^{-e}}}
\providecommand{\nonbest}[1]{\ensuremath{#1^{-n}}}
\providecommand{\p}{\ensuremath{p}}
\providecommand{\pset}{\ensuremath{\mathbf{p}}}
\providecommand{\s}{\ensuremath{s}}
\providecommand{\sset}{\ensuremath{\mathbf{s}}}
\providecommand{\nsset}{\ensuremath{\mathbf{ns}}}
\providecommand{\n}{\ensuremath{n}}
\providecommand{\nset}{\ensuremath{\mathbf{n}}}
\providecommand{\nnset}{\ensuremath{\mathbf{nn}}}

@@

To enumerate all of the possible next positions of a particle, one must simply
evaluate the motion equations for all possible combinations of personal best
and neighborhood best updates that a particle could see.  This produces
$2(n+1)$ speculative positions to be evaluated, where $n$ is the number of
neighbors the particle has.  If a particle is also its own neighbor, as is most
often the case, one of the $2(n+1)$ speculative positions can be eliminated, as
a particle cannot be the source of its $\nbest$ update while not updating its
$\pbest$.  Table~\ref{tab:evals} lists all possible combinations of updates
that can be seen by a particle with two neighbors other than itself, which is
the case in all of the topologies we use.  If all of these speculative
positions are evaluated in parallel along with the original, non-speculative
position of the particle at time $t-1$, then one of the speculative positions
has to be the correct one, and two iterations have been evaluated at once.

\begin{table}
  \caption{All possible updates for a particle with two neighbors.}
  \label{tab:evals}
  \centering
  \begin{tabular}{c|c}
	$\pbest$ updated&Source of $\nbest$ update\\
	\hline
	No update&No update\\
	Updated&No update\\
	No update&Neighbor 1\\
	No update&Neighbor 2\\
	Updated&The particle itself\\
	Updated&Neigbhor 1\\
	Updated&Neighbor 2\\
  \end{tabular}
\end{table}

@@

To aid in describing our methodology, we introduce a few terms.  A particle's
set of \emph{speculative children} is the set of all possible next iteration
states that a particle could have.  We use $\p_t$ to denote a particle at
iteration $t$ and $\s_{t+1}$ to denote one of $\p_t$'s speculative children,
corresponding to one of the rows in Table~\ref{tab:evals}.  $\n_t$ is a
neighbor of particle $\p_t$.  A set of particles is $\pset$, $\sset$, or
$\nset$, whereas a single particle is simply $\p$, $\s$, or $\n$.

@@

Given a set of particles at iteration $t-1$ (perhaps which have just been
initialized), the master must move each particle using
Equations~\eqref{eq:velupdate} and \eqref{eq:posupdate} to obtain the set
$\noeval{\pset}_t$.  For each particle $\noeval{\p}_t$, the master must get its
set of neighbors $\noeval{\nset}_t$ and use their positions, along with the
position of $\noeval{\p}_t$, to enumerate all possible combinations of $\pbest$
and $\nbest$ that particle $\noeval{\p}_t$ could end up with (as shown in
Table~\ref{tab:evals}).  Each of those combinations defines a speculative
particle, $\s_t$, that can be moved using the motion equations to obtain
$\noeval{\s}_{t+1}$.  The result is a set of particles $\noeval{\pset}_t$, and
for each particle a set of speculative children $\noeval{\sset}_{t+1}$, which
can all be evaluated at once.

@@

To get that information, a round of message passing is required.  Each particle
sends its position to its neighbors at iteration $t$, so that all particles can
generate $\noeval{\sset}_{t+1}$.  After each particle evaluates its position
and the positions of its speculative children, it passes information about the
outcome iteration $t$ to its neighbors, so that each particle can receive
information about its neighborhood best to be updated to $\p_t$.  Once that
communication is finished, the particle can select the speculative child which
matched the branch that iteration $t$ actually produced.  Then another round of
information passing follows, for iteration $t+1$, so that $\nonbest{\p}_{t+1}$
can be updated to $\p_{t+1}$.  Two iterations have then been completed with
only one round of evaluations, and the next iteration can start again with the
first round of message passing.

\subsection{Alternate form of message passing}

In some distributed frameworks, like the MapReduce framework that we used in
our experiments, rounds of message passing can be expensive.  The method
just described uses three rounds of message passing for every two iterations.
Accordingly, we describe a method that requires only one round of communication
for each pair of iterations.  A lot more messages are needed, but that is
sometimes more desirable than synchronizing all of the machines three times.

@@

Upon evaluation, $\noeval{\p}_t$ becomes $\nonbest{\p}_t$, needing only to get
its neighborhood best information from its neighbors.  All of its neighbors,
then, must send it a message, so that from their updated personal best at
iteration $t$ the particle becomes $\p_t$.  The work done with the messages
received thus far is just as in regular PSO, and is graphically depicted in 
\fig{messages1}.

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \includegraphics[width=.3\columnwidth]{messages1.eps}
  \caption{The messages needed to produce $\p_{t}$.  Items without an arrow
  pointing to them are messages the particle needs to receive.}
  \label{fig:messages1}
\end{figure}

With $\p_t$ we can select the correct speculative child as described above and
produce $\nonbest{\p}_{t+1}$.  Again we show the use of messages thus far
graphically, in \fig{messages2}.

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \includegraphics[width=.5\columnwidth]{messages2.eps}
  \caption{The messages needed to produce $\nonbest{\p}_{t+1}$.  Items without
  an arrow pointing to them are messages the particle needs to receive.}
  \label{fig:messages2}
\end{figure}

We then need the set of neighbors to $\nonbest{\p}_{t+1}$,
$\nonbest{\nset}_{t+1}$, so we can update $\nonbest{\p}_{t+1}$'s neighborhood
best.  To produce each neighbor $\nonbest{\n}_{t+1}$, we need the same
information for the neighboring particle that we needed to produce the original
particle, $\nonbest{\p}_{t+1}$; we need the original neighbor particle, its
speculative children, and its neighbors.  With that information, the set
$\nonbest{\nset}_{t+1}$ can be obtained by following the same process used to
obtain $\nonbest{\p}_{t+1}$.  We graphically show the messages needed to
produce $\nonbest{\p}_{t+1}$ in \fig{messages3}.  Note that it looks identical
to \fig{messages2}, just with different sets of particles.

\begin{figure}
  \centering
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \includegraphics[width=.6\columnwidth]{messages3.eps}
  \caption{The messages needed to produce $\nonbest{\nset}_{t+1}$.  Items
  without an arrow pointing to them are messages the particle needs to receive.
  $\nnset$ is the set of neighbors for each particle $\n$, and $\nsset$ is the
  set of $\n$'s speculative children.  Note the similarity between this and
  \fig{messages2}.}
  \label{fig:messages3}
\end{figure}

With $\nonbest{\nset}_{t+1}$ and $\nonbest{\p}_{t+1}$, we can produce
$\p_{t+1}$.  This is shown in \fig{messages4}.  Note that we just combined
Figures~\ref{fig:messages2}~and~\ref{fig:messages3}, putting them together
to make $\p_{t+1}$, as all the particle needs is it's neighborhood best to be
updated.

\begin{figure}
  \centering
  \psfrag{pt-n}[C][C]{$\nonbest{\p}_{t}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{pt}[C][C]{$\p_{t}$}
  \psfrag{pt1-n}[C][C]{$\nonbest{\p}_{t+1}$}
  \psfrag{st1-n}[C][C]{$\nonbest{\sset}_{t+1}$}
  \psfrag{nt-n}[C][C]{$\nonbest{\nset}_{t}$}
  \psfrag{nnt-n}[C][C]{$\nonbest{\nnset}_{t}$}
  \psfrag{nt}[C][C]{$\nset_{t}$}
  \psfrag{nt1-n}[C][C]{$\nonbest{\nset}_{t+1}$}
  \psfrag{nst1-n}[C][C]{$\nonbest{\nsset}_{t+1}$}
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \includegraphics[width=.8\columnwidth]{messages4.eps}
  \caption{The messages needed to produce $\p_{t+1}$.  Items
  without an arrow pointing to them are messages the particle needs to receive.
  Note that this is just a combination of \fig{messages2} and \fig{messages3}.}
  \label{fig:messages4}
\end{figure}

In order to get $\p_{t+1}$, then, a particle needs to receive messages from its
neighbors, its neighbors' neighbors, its speculative children, and its
neighbors' speculative children.  The particle $\p_{t+1}$ can be passed to some
central machine to track the progress of the algorithm, and it can be moved to
$\noeval{\p}_{t+2}$ in order to start the next iteration.

@@

We have already described how to use a set of messages to obtain $\p_{t+1}$.
The process is exactly the same to produce each $\n_{t+1}$, requiring the same
messages, except for the neighbor particles instead of the particle itself.  
\fig{messages5} shows graphically how $\noeval{\sset}_{t+3}$ is produced.

\begin{figure}
  \centering
  \psfrag{pt1}[C][C]{$\p_{t+1}$}
  \psfrag{nt1}[C][C]{$\nset_{t+1}$}
  \psfrag{nt2-e}[C][C]{$\noeval{\nset}_{t+2}$}
  \psfrag{pt2-e}[C][C]{$\noeval{\p}_{t+2}$}
  \psfrag{st3-e}[C][C]{$\noeval{\sset}_{t+3}$}
  \includegraphics[width=.5\columnwidth]{messages5.eps}
  \caption{The messages needed to produce $\noeval{\sset}_{t+3}$. $\p_{t+1}$
  and $\nset_{t+1}$ are both produced as in \fig{messages4}.}
  \label{fig:messages5}
\end{figure}

@@

This may seem like an inordinate amount of work, and in some instances it is.
However, some parallel frameworks necessitate this type of message passing, so
we showed how speculative evaluation can be performed in such frameworks.

@@

Performing speculative evaluation in PSO with a dynamic topology raises a few
issues of its own.  First, with neighbors changing each iteration, messages
that processors pass to their neighbors need to be sent to the correct
neighbors for each iteration.  In centralized algorithms and in distributed
algorithms with several rounds of message passing (the first method described
above), nothing needs to be changed.  However, in the case where processors
reconstruct information about neighboring particles from messages, those
messages need to be sent to the correct neighbors.  A particle cannot simply
send messages to its neighbors' neighbors' neighbors---it needs to send
messages to its iteration $t$ neighbors' iteration $t+1$ neighbors, and so on.
For every neighbor outward information is sent, the iteration also needs to be
incremented, as information about neighbors' neighbors is used during iteration
$t+1$, and information about neighbors' neighbors' neighbors is used to
reconstruct information about iteration $t+2$.  Also, this method of message
passing again requires the use of random seeds if the topology is random.

@@

This can be thought of as drawing a number of samples from the next iteration
and accepting the best one.  Speculative particles that move in good directions
are kept.  It seems this modification to PSO would produce more exploitation,
decreasing the amount of exploration.  With functions that are less deceptive
we would expect this method to work well, while performance on more deceptive
functions would probably be hurt.

@@

In our experiments we compared our speculative PSO algorithm to the standard
PSO algorithm.  To use the same number of processors for each algorithm,
speculative evaluation needs a smaller swarm size than standard PSO.  For each
of the topologies we used, a particle has two neighbors other than itself.  As
shown in Table~\ref{tab:evals}, this results in $7$ speculative evaluations per
particle.  With one evaluation needed for the original, non-speculative
particle, we have $8p_s$ evaluations for every two iterations, where $p_s$ is
the number of particles in the speculative swarm.  The extra evaluations done
in speculative evaluation would instead be particles in regular PSO, so we
compare swarms of size $p_s$ in speculative evaluation with swarms of size
$8p_s$ in regular PSO.

We use a Random topology in our speculative algorithm, as the Complete topology
leads to an explosion of speculative evaluations.  However, if speculative
evaluation were not being performed, it is likely that the Complete topology
would be used, ignoring the communication cost of Complete over Random.  To be
fair in our comparisons, for the standard PSO algorithm we compare to both the
Random topology with two neighbors and to the Complete topology.  All functions
have 20 dimensions, and all results shown are averages over 20 runs.  We use
240 processors in our experiments, so speculative algorithms use 30 particles
and standard PSO uses 240 particles.  For ease of labeling, we call our
speculative algorithm SEPSO, the variant SEPSO Pick Best, and standard PSO just
PSO.

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
small Complete swarm in \fig{sphere1}.  In comparing SEPSO with PSO, we can see
that SEPSO clearly beats PSO with a Random topology.  However, the performance
between SEPSO and PSO Complete is not as encouraging---PSO Complete performs
slightly better.  But SEPSO Pick Best handily outperforms all other algorithms
in this problem.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere1.eps}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:sphere1}
\end{figure}

Next we look at the function Griewank, defined by the equation $f(\Vec{x}) =
\frac{1}{4000} \sum_{i=1}^D x_i^2 - \Pi_{i=1}^D \cos\left(\frac{x_i}{\sqrt{i}}
\right) + 1$.  It is generally recommended to use the Ring topology when
optimizing the Griewank function, as Complete is prone to premature convergence
on a local optima.  We show results in \fig{griewank1} for swarms of size 30
and 240 using the Ring topology. One can see in the figure that both SEPSO
algorithms initially outperformed PSO, but got stuck in a local optima and were
eventually outperformed.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Speculative algorithms have a swarm of 30 particles and do two
  iterations per round, while standard PSO algorithms use 240 processors and do
  one iteration per round.}
  \label{fig:griewank1}
\end{figure}

\fig{griewank1} is slightly deceptive, however, as with Griewank some of the
runs found the global optimum, 0, while some did not.  A more accurate plot for
this function is one showing what percentage of the runs have found the global
optimum at each iteration.  \fig{griewank2} shows such a plot.  A swarm of size
30 performing the PSO algorithm gets stuck a little less than half of the time,
while 240 particles provides enough exploration to find the global optimum
every time.  Surprisingly, using the SEPSO Pick Best algorithm with 30
particles greatly improves the chance of success with Griewank, while also
finding the global optimum in far fewer rounds of evaluations than standard
PSO.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2.eps}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that found the global optimum at each iteration.}
  \label{fig:griewank2}
\end{figure}

Lastly we look at a function that does not show much or any improvement with
this method.  The function Rastrigin is defined as $f(\Vec{x}) = \sum_{i=1}^D
\left(x_i^2 - 10\cos\left(2\pi x_i\right) + 10\right)$.  It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000 particles~\cite{mcnabb-cec09}.  Smaller
swarms get caught in local optima.  Because our speculative algorithms have a
significantly smaller swarm size, they get stuck at higher values while the
larger swarms performing regular PSO continue to improve the best value found.
\fig{rastrigin} shows the results graphically.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin.eps}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  Speculative algorithms have a swarm of 30 particles and do
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:rastrigin}
\end{figure}

@@

We have described how parallel implementations of particle swarm optimization
can be modified to allow additional processors to increase the number of
iterations of the algorithm performed, instead of merely adding more particles
to the swarm.  We have detailed several different implementations of the
algorithm, depending on the parallel PSO framework being used.  Using our
modifications, the original PSO algorithm is exactly reproduced two iterations
at a time.  This technique requires more function evaluations per iteration
than regular PSO, but for some functions still performs better when run in a
parallel environment.

We have also described a method for making use of extra speculative evaluations
that performs incredibly well on some functions.

@@

There are other functions for which there is a point where the algorithm has
done ``enough'' exploration and consistently finds the global optimum.  Adding
additional particles to the swarm is akin to increasing its capacity for
exploration.  We have seen that for a 20-dimensional Griewank function, a
number of particles greater than 30 but less than 240 has ``enough''
exploration and finds the optimal value every time.  For these functions, one
should use processors to increase the swarm size until enough exploration has
been reached, then use additional processors to increase the number of
iterations performed.
